---
title: "Antibody Visualization"
author:
- name: Nick Borcherding
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
package: deepMatchR
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Antibody Visualization}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8
)
library(BiocStyle)
```

# Introduction

Visualizing HLA antibody data is essential for:

- Monitoring patient sensitization
- Identifying unacceptable antigens for transplant
- Tracking antibody trends over time
- Analyzing eplet-level reactivity patterns

This article covers **deepMatchR**'s visualization tools for Single Antigen Bead (SAB) and Panel Reactive Antibody (PRA) assay results.

# Loading the Package

```{r load-package}
library(deepMatchR)
library(dplyr)
library(ggplot2)
```

# Example Data

**deepMatchR** includes simulated assay data for demonstrations:

```{r load-data}
data("deepMatchR_example")

# Available datasets:
# 1. Class I SAB
# 2. Class II SAB
# 3. PRA

head(deepMatchR_example[[1]], 5)
```

## Data Format Requirements

For downstream analysis, your data must include these columns:

| Column | Description | Example |
|--------|-------------|---------|
| `BeadID` | Unique bead identifier | 1, 2, 3... |
| `SpecAbbr` | Abbreviated specificity | A01, B35 |
| `Specificity` | Full specificity string | "A*01:01, B*07:01" |
| `NormalValue` | Measured MFI value | 15234 |

# Plotting Antibodies

## SAB Bar Plot

The `plotAntibodies()` function creates publication-quality visualizations:

```{r plot-sab}
plotAntibodies(
  result_file = deepMatchR_example[[1]],
  type = "SAB",
  bead_cutoffs = c(2000, 1000, 500, 250),
  add_table = TRUE,
  palette = "spectral"
)
```

## Highlighting Specific Antigens

Focus on antigens of interest using `highlight_antigen`:

```{r plot-sab-highlight}
plotAntibodies(
  result_file = deepMatchR_example[[1]],
  type = "SAB",
  bead_cutoffs = c(2000, 1000, 500),
  add_table = TRUE,
  palette = "spectral",
  highlight_antigen = c("Bw4", "Bw6")
)
```

## PRA Bar Plot

```{r plot-pra, fig.height=6}
plotAntibodies(
  result_file = deepMatchR_example[[3]],
  type = "PRA",
  class = "I",
  add_table = TRUE,
  palette = "spectral"
)
```

## Time-Series Trend Plot

Track antibody levels over multiple time points:

```{r plot-trend, fig.height=6}
# Simulate longitudinal data
set.seed(123)
sab_data_list <- list(
  "01/01/2023" = deepMatchR_example[[1]],
  "02/01/2023" = deepMatchR_example[[1]] %>%
    mutate(NormalValue = NormalValue * runif(n(), 0.5, 0.8)),
  "03/01/2023" = deepMatchR_example[[1]] %>%
    mutate(NormalValue = NormalValue * runif(n(), 1, 3))
)

plotAntibodies(
  result_file = sab_data_list,
  type = "SAB",
  plot_trend = TRUE,
  highlight_threshold = 5000,
  vline_dates = c("2023-02-15")
)
```

# Eplet Visualization

## `plotEplets()` Overview

The `plotEplets()` function offers three visualization types:

| Type | Best For |
|------|----------|
| `treemap` | Overview of eplet prominence |
| `bar` | Ranking eplets by positivity |
| `AUC` | Performance across thresholds |

## Treemap View

Tile area represents importance (positive beads × positivity rate):

```{r plot-treemap}
plotEplets(
  result_file = deepMatchR_example[[1]],
  plot_type = "treemap",
  cutoff = 2000,
  evidence_level = c("A1", "A2", "B"),
  percPos_filter = 0.4,
  palette = "spectral"
)
```

## Bar Plot

Ranks eplets by positivity rate at a specific threshold:

```{r plot-bar, fig.height=6}
plotEplets(
  result_file = deepMatchR_example[[1]],
  plot_type = "bar",
  group_by = "loci",
  cutoff = 2000,
  evidence_level = c("A1", "A2", "B"),
  percPos_filter = 0.4,
  top_eplets = 20,
  palette = "spectral"
)
```

## AUC Bar Plot

Ranks eplets by performance across multiple thresholds:

```{r plot-auc-bar, fig.height=6}
plotEplets(
  result_file = deepMatchR_example[[1]],
  plot_type = "AUC",
  percPos_filter = 0.4,
  group_by = "evidence_level",
  cut_min = 250,
  cut_max = 10000,
  cut_step = 250,
  top_eplets = 20,
  palette = "spectral"
)
```

# Eplet AUC Analysis

## `epletAUC()` Function

Calculates the Area Under the Curve for eplet reactivity across multiple MFI thresholds. This provides a more robust measure than a single cutoff.

## Visualize Reactivity Curves

```{r eplet-auc-plot, fig.height=6}
epletAUC(
  result_file = deepMatchR_example[[1]],
  group_by = "evidence",
  evidence_level = c("A1", "A2", "B", "D"),
  plot_results = TRUE,
  cut_min = 250,
  cut_max = 10000,
  cut_step = 250,
  palette = "inferno"
)
```

## Get AUC Values as Data

```{r eplet-auc-data}
auc_result <- epletAUC(
  result_file = deepMatchR_example[[1]],
  plot_results = FALSE,
  cut_min = 250,
  cut_max = 10000,
  cut_step = 250
)
head(auc_result)
```

## Advanced Filtering

```{r eplet-auc-filter, fig.height=6}
epletAUC(
  result_file = deepMatchR_example[[1]],
  label = FALSE,
  plot_results = TRUE,
  eplet_filter = 10,      # Minimum bead count
  percPos_filter = 1.0    # Minimum positivity threshold
)
```

# Customization Options

## Color Palettes

Available palettes include:

- `spectral` - Rainbow diverging
- `inferno` - Yellow to purple
- `viridis` - Green to purple
- `plasma` - Yellow to magenta

```{r palette-demo, fig.height=6}
# Using inferno palette
plotEplets(
  result_file = deepMatchR_example[[1]],
  plot_type = "bar",
  cutoff = 2000,
  top_eplets = 15,
  palette = "inferno"
)
```

## Table Options

Control the summary table with `add_table`:

```{r table-options, fig.height=6}
# Without table
plotAntibodies(
  result_file = deepMatchR_example[[1]],
  type = "SAB",
  add_table = FALSE
)
```

# Function Reference

## `plotAntibodies()`

| Argument | Default | Description |
|----------|---------|-------------|
| `result_file` | -- | SAB/PRA data frame or named list for trends |
| `type` | `"SAB"` | Assay type: "SAB" or "PRA" |
| `class` | `NULL` | HLA class filter: "I" or "II" |
| `bead_cutoffs` | `c(2000,1000,500,250)` | MFI cutoff thresholds |
| `add_table` | `TRUE` | Include summary table |
| `palette` | `"spectral"` | Color palette |
| `highlight_antigen` | `NULL` | Antigens to highlight |
| `plot_trend` | `FALSE` | Create time-series plot |
| `vline_dates` | `NULL` | Vertical line dates for trends |

## `plotEplets()`

| Argument | Default | Description |
|----------|---------|-------------|
| `result_file` | -- | SAB data frame |
| `plot_type` | `"treemap"` | "treemap", "bar", or "AUC" |
| `cutoff` | `2000` | MFI cutoff for positive |
| `evidence_level` | `c("A1","A2","B")` | Evidence levels to include |
| `percPos_filter` | `0.4` | Minimum positivity rate |
| `top_eplets` | `20` | Number of top eplets for bar plots |
| `group_by` | `"evidence"` | Grouping variable |
| `palette` | `"spectral"` | Color palette |

## `epletAUC()`

| Argument | Default | Description |
|----------|---------|-------------|
| `result_file` | -- | SAB data frame |
| `plot_results` | `TRUE` | Generate plot |
| `cut_min`, `cut_max` | `250`, `10000` | MFI range |
| `cut_step` | `250` | Step size |
| `eplet_filter` | `NULL` | Minimum bead count |
| `percPos_filter` | `NULL` | Minimum final positivity |
| `label` | `TRUE` | Add labels to plot |
| `palette` | `"spectral"` | Color palette |

# Session Information

```{r sessioninfo}
sessionInfo()
```
