title: Using deepMatchR in research and clinical workflows
author: 
- name: Nick Borcherding
  email: ncborch@gmail.com
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA

date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'

output:
  BiocStyle::html_document:
    toc_float: true
package: deepMatchR
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Using deepMatchR}
  %\VignetteEncoding{UTF-8} 
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Introduction

Human leukocyte antigen (HLA) are a central node in innate and adaptive immune responses. The characterization of HLA antibodies and typing play a pivotal role in bone marrow and organ transplantation. **deepMatchR** is a collection of tools using advanced statistical approaches and deep learning to better characterize HLA immunogenicity in the context of clinical or research questions. 

More information is available at the [deepMatchR GitHub Repo](https://github.com/BorchLab/deepMatchR).

## Loading Libraries

```{r}
suppressMessages(library(deepMatchR))
```

## Loading Built-In Data

**deepMatchR** has a set of 2 simulated class I and class II SAB assay results with clear patterns of reactivity and eplets. We will explore the analysis of these data sets further in later sections. It is important to note that both examples have fixed formatting that is important to follow. 

Specifically required for downstream analysis: 

* `BeadID`: Identifies each bead in the assay
* `SpecAbbr`: Abbreviated specificity (e.g., A01, B35, etc.)
* `Specifcity`: Full specificity strings (e.g., "A\*01:01, B\*07:01")
* `NormalValue`: The measured MFI for each bead

```{r}
data("deepMatchR_example")

#SAB Class I example Data
head(deepMatchR_example[[1]])

#SAB Class II example Data
head(deepMatchR_example[[2]])
```

## plotSAB 

The ```plotSAB()``` function is designed to simplify the visualization of SAB data. It accepts either a data frame of SAB results or a file path, processes the data by cleaning, categorizing MFI values based on provided bead cutoff values, and applies a customizable color palette. Optionally, you can append an antigen-level table to the plot for further detail, with the ability to highlight specific antigens.

Below is an example of how to use plotSAB with the deepMatchR_example[[1]] object. In this example, we set custom bead cutoffs and highlight antigens "A1" and "B8".

```{r}
# Generate the SAB plot with antigen-level details and highlight specific antigens
plotSAB(result_file = deepMatchR_example[[1]],
             bead_cutoffs = c(2000, 1000, 500, 250),
             add_table = TRUE,
             palette = "spectral",
             highlight_antigen = c("Bw4"))
```

### Interpretation

* **Bar Plot**: The main plot displays the MFI values of each bead, categorized into various levels. The color of each bar corresponds to its MFI category, helping you quickly identify beads that are above or below defined thresholds.

* **Antigen-Level Table**: When `add_table` is set to TRUE, a table is added below the main plot. This table lists the loci and antigen information for each bead. Antigens specified in the `highlight_antigen` parameter (here, "Bw4") are highlighted in red, making it easier to identify them within the dataset.

## epletAUC

The `epletAUC()` function provides a flexible way to assess eplet-level reactivity across a range of MFI cutoffs in SAB assays. 

Basic Workflow:

* Takes SAB data (either from a data frame or file path)
* Identifies eplets present on the tested alleles
* Assesses positivity over a range of MFI cutoff values
* Optionally filters out low-frequency or lower-confidence eplets
* Calculates the area under the curve (AUC) for each eplet’s positivity profile

You can either plot these positivity curves or return a summarized table of AUC values, depending on your analysis needs.

```{r}
plot_result <- epletAUC(
  result_file   = deepMatchR_example[[1]],
  plot_results  = TRUE,         # Request a plot
  cut_min       = 250,          # Starting cutoff
  cut_max       = 10000,        # Ending cutoff
  cut_step      = 250,          # Step between cutoffs
  palette       = "inferno"     # Color palette for lines
)

plot_result
```
If you prefer a data table with the computed AUC values, simply set plot_results = FALSE. The function will then compute and return a tidy data frame (tibble):

```{r}
auc_result <- epletAUC(
  result_file   = deepMatchR_example[[1]],
  plot_results  = FALSE,  # Return AUC data instead of a plot
  cut_min       = 250,
  cut_max       = 10000,
  cut_step      = 250
)

auc_result
```

We can use more advanced filtering strategies to extract clinically actionable data. `eplet_filter` allows us to consider eplets that appear at least a certain number of times in your assay. Likewise, `percPos_filter` can be used to threshold the maximumn positivity for each eplet across the cutoffs.


```{r}
epletAUC(
  result_file   = deepMatchR_example[[1]],
  label = FALSE,
  plot_results  = TRUE,
  eplet_filter  = 10,
  percPos_filter = 1.0
)
```
## plotEplets

The ```plotEplets()``` function is designed to facilitate the visualization of eplet-specific assay results from single antigen bead (SAB) experiments. It streamlines the process of reading SAB data, merging it with an internal eplet annotation database, and then generating one of three different plot types: a treemap, a bar plot, or an AUC plot. Each plot type highlights different aspects of the data:

* **Treemap**: Displays eplet counts with areas proportional to the product of the positive bead count and the positive bead percentage.
* **Bar Plot**: Ranks eplets based on the proportion of positive beads, providing a clear ranking of top eplets.
* **AUC** Plot: Ranks eplets by their normalized AUC across a range of MFI cutoffs, offering a dynamic view of eplet performance over the full assay range.

### Generating a Treemap

In the treemap visualization, the area of each rectangle represents the product of the number of positive beads and the positive bead percentage for each eplet. This plot is ideal for getting an overall sense of eplet performance across the assay.

```{r}
plotEplets(result_file = deepMatchR_example[[1]],
           plot.type = "treemap",
           cutoff = 2000,
           evidence_level = c("A1", "A2", "B"),
           percPos_filter = 0.4,
           palette = "spectral")
```

### Generating a Barplot

The bar plot ranks eplets by the proportion of positive beads defined by the cutoff. This visualization is useful when you want to focus on the top-performing eplets based on their positivity rate

```{r}
plotEplets(result_file = deepMatchR_example[[1]],
           plot.type = "bar",
           cutoff = 2000,
           evidence_level = c("A1", "A2", "B"),
           percPos_filter = 0.4,
           top.eplets = 20,
           palette = "spectral")
```

### Generating AUC Barplot

The AUC plot visualizes the normalized area under the curve for eplets across a range of MFI cutoffs. This approach provides insight into the performance consistency of each eplet across different assay thresholds.

```{r}
plotEplets(result_file = deepMatchR_example[[1]],
           plot.type = "AUC",
           percPos_filter = 0.4,
           cut_min = 250,
           cut_max = 10000,
           cut_step = 250,
           top.eplets = 20,
           palette = "inferno")
```


***

# Conclusions

If you have any questions, comments or suggestions, feel free to visit the [github repository](https://github.com/ncborcherding/deepMatchR).

```{r}
sessionInfo()
```