---
title: Using deepMatchR in research and clinical workflows
author: 
- name: Nick Borcherding
  email: ncborch@gmail.com
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA

date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'

output:
  BiocStyle::html_document:
    toc_float: true
package: deepMatchR
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Using deepMatchR}
  %\VignetteEncoding{UTF-8} 
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Introduction

Human leukocyte antigen (HLA) are a central node in innate and adaptive immune responses. The characterization of HLA antibodies and typing play a pivotal role in bone marrow and organ transplantation. **deepMatchR** is a collection of tools using advanced statistical approaches and deep learning to better characterize HLA immunogenicity in the context of clinical or research questions. 

More information is available at the [deepMatchR GitHub Repo](https://github.com/BorchLab/deepMatchR).

## Loading Libraries

```{r}
library(deepMatchR)
library(dplyr)
```

## Loading Built-In Data

**deepMatchR** has a set of 3 simulated assay outputs, 1) class I SAB, 2) class II SAB, and 3) PRA assay results with clear patterns of reactivity and eplets. We will explore the analysis of these data sets further in later sections. It is important to note that that the examples have fixed formatting.

Specifically required for downstream analysis: 

* `BeadID`: Identifies each bead in the assay
* `SpecAbbr`: Abbreviated specificity (e.g., A01, B35, etc.)
* `Specifcity`: Full specificity strings (e.g., "A\*01:01, B\*07:01")
* `NormalValue`: The measured MFI for each bead

```{r}
data("deepMatchR_example")

#SAB Class I example data
head(deepMatchR_example[[1]])

#SAB Class II example data
head(deepMatchR_example[[2]])

#PRA example data
head(deepMatchR_example[[3]])
```

# Sequence-Level Analysis

`deepMatchR` includes several functions for comparing and analyzing HLA sequences at the amino acid and eplet level. These are fundamental tools for quantifying the differences between alleles.

## Comparing Sequences with `compareHlaSequences`

The `compareHlaSequences` function allows you to see the specific amino acid differences between two protein sequences. It not only identifies the positions of mismatches but also characterizes them based on changes in physicochemical properties (polarity and charge), which can be important for predicting immunogenicity.

**Arguments:**

*   `seq1`, `seq2`: The two protein sequences to compare. They must be of the same length.

**Example:**

Here, we compare two sequences that differ at two positions. The output is a data frame detailing each mismatch.

```{r}
seq1 <- "YFAMYGEKVAHTHVDTLYVRYHY"
seq2 <- "YFDMYGEKVAHTHVDTLYVRFHY"
compareHlaSequences(seq1, seq2)
```

## Quantifying Mismatches with `quantifyMismatch`

If you only need to know the total number of differences, `quantifyMismatch` provides a quick and simple way to get a single integer value representing the mismatch count.

**Arguments:**

*   `seq1`, `seq2`: The two protein sequences to compare.

**Example:**

```{r}
quantifyMismatch(seq1, seq2)
```

## Retrieving Allele Sequences with `getAlleleSequence`

To perform these comparisons, you first need the protein sequences for the HLA alleles of interest. The `getAlleleSequence` function retrieves these sequences from the IMGT/HLA database. Note that this function requires an internet connection.

**Arguments:**

*   `allele_name`: The name of the HLA allele (e.g., `"A*01:01"`).

**Example:**

```{r, results='hide'}
# Requires internet connection
a0101_seq <- getAlleleSequence("A*01:01")
head(a0101_seq)
```

## Quantifying Eplet Mismatches with `quantifyEpletMismatch`

Eplets are key components of HLA immunogenicity. The `quantifyEpletMismatch` function calculates the number of eplets that are not shared between two HLA alleles, giving a measure of the potential for an antibody response. It uses an internal database of known eplets.

**Arguments:**

*   `allele1`, `allele2`: The names of the two HLA alleles to compare.
*   `evidence_level`: Filters eplets based on their level of experimental evidence (e.g., `"A1"`, `"A2"`, `"B"`).

**Example:**

```{r}
quantifyEpletMismatch("A*01:01", "A*02:01")
```

### Calculating Eplet and Mismatch Load

When working with donor and recipient HLA genotypes, it is often useful to calculate the total eplet or amino acid mismatch load. The `calculateEpletLoad` and `calculateMismatchLoad` functions provide a streamlined way to do this.

**Arguments:**

*   `recipient_geno`, `donor_geno`: `hla_genotype` objects for the recipient and donor.
*   `evidence_level` (for `calculateEpletLoad`): Filters eplets based on their level of experimental evidence.

**Example:**

First, we create `hla_genotype` objects for a recipient and a donor.

```{r}
# Create dummy recipient and donor genotypes
recipient <- data.frame(
  A_1 = "A*01:01", A_2 = "A*02:01",
  B_1 = "B*07:02", B_2 = "B*08:01"
)
donor <- data.frame(
  A_1 = "A*03:01", A_2 = "A*24:02",
  B_1 = "B*44:02", B_2 = "B*51:01"
)
# Convert to hla_genotype objects
recipient_geno <- hlaGeno(recipient)
donor_geno <- hlaGeno(donor)
```

Now we can use these objects to calculate the eplet and mismatch loads.

```{r}
# Calculate eplet load
calculateEpletLoad(recipient_geno, donor_geno)

# Calculate mismatch load
donor_geno2 <- hlaGeno(data.frame(
  A_1 = "A*01:01", A_2 = "A*03:01",
  B_1 = "B*44:02", B_2 = "B*51:01"
))
calculateMismatchLoad(recipient_geno, donor_geno2)
```

## plotAntibodies

The `plotAntibodies()` function is a versatile tool for visualizing antibody data from both Single Antigen Bead (SAB) and Panel Reactive Antibody (PRA) assays. Its primary goal is to create clear, publication-quality bar plots of Mean Fluorescence Intensity (MFI) values. For added detail, it can append a table of antigen specificities directly below the plot, which is useful for interpreting complex results. The function is flexible, accepting data as either a data frame or directly from a file.

**Key Arguments:**

*   `result_file`: The input data, which can be a data frame or a path to a CSV/XLS file.
*   `type`: Must be either `"SAB"` or `"PRA"` to specify the assay type.
*   `add_table`: A logical value (`TRUE`/`FALSE`) to control whether the antigen table is displayed.
*   `highlight_antigen`: Allows you to specify one or more antigens to highlight in red for emphasis.

### Bar Plot for SAB data

Here is an example of how to use `plotAntibodies` for SAB data. We set `type = "SAB"` and highlight the `"Bw4"` antigen.

```{r}
plotAntibodies(result_file = deepMatchR_example[[1]],
               type = "SAB",
               bead_cutoffs = c(2000, 1000, 500, 250),
               add_table = TRUE,
               palette = "spectral",
               highlight_antigen = c("Bw4"))
```

### Bar Plot for PRA data

For PRA data, we set `type = "PRA"` and specify the `class` ("I" or "II").

```{r}
plotAntibodies(result_file = deepMatchR_example[[3]],
               type = "PRA",
               class = "I",
               add_table = TRUE,
               palette = "spectral")
```

### Time-Series Trend Plot

The function can also plot MFI values over time. For this, `plot_trend` must be set to `TRUE`, and `result_file` should be a named list of data frames, where the names are the dates of the assays.

Here is an example with simulated data:

```{r}
# Simulate a list of SAB results over time
set.seed(123)
sab_data_list <- list(
  "01/01/2023" = deepMatchR_example[[1]],
  "02/01/2023" = deepMatchR_example[[1]] %>% mutate(NormalValue = NormalValue * runif(n(), 0.5, 0.8)),
  "03/01/2023" = deepMatchR_example[[1]] %>% mutate(NormalValue = NormalValue * runif(n(), 1, 3))
)

plotAntibodies(result_file = sab_data_list,
               type = "SAB",
               plot_trend = TRUE,
               highlight_threshold = 5000,
               vline_dates = c("2023-02-15"))
```

## spiDeconvolute

The `spiDeconvolute()` function addresses a common challenge in HLA labs: reconciling results from Single Antigen Bead (SAB) and Panel Reactive Antibody (PRA) assays. It systematically compares the two, identifying which antibody reactivities are supported by both assays (concordant) and which are found in only one (discordant). This is crucial for distinguishing true reactivities from potential artifacts.

The function returns a list containing four data frames: `concordant`, `sabOnly`, `praOnly`, and a `summary` of the deconvolution process.

**Key Arguments:**

*   `SAB`: A data frame of your SAB results, expected to have `antigen` and `NormalValue` columns.
*   `PRA`: A data frame of your PRA results, with `BeadID` and `NormalValue`.
*   `panelLong`: A data frame mapping the antigens on each bead of the PRA panel, with `BeadID` and `antigen` columns.

Here is an example of how to use `spiDeconvolute()` with sample data:

```{r}
# 1. Create sample data
sab_data <- data.table::data.table(
  antigen = c("A1", "A2", "B7", "B8", "DR1"),
  NormalValue = c(2000, 500, 3000, 400, 4000),
  class = c("I", "I", "I", "I", "II")
)
pra_data <- data.table::data.table(
  BeadID = c(1, 2, 3, 4),
  NormalValue = c(1500, 400, 2000, 3000)
)
panel_data <- data.table::data.table(
  BeadID = c(1, 1, 2, 3, 3, 4),
  antigen = c("A1", "B8", "A2", "B7", "B8", "DR1")
)

# 2. Run the deconvolution
decon_results <- spiDeconvolute(
  SAB = sab_data,
  PRA = pra_data,
  panelLong = panel_data
)

# 3. View the results
# Concordant antigens
print(decon_results$concordant)

# SAB-only antigens
print(decon_results$sabOnly)

# PRA-only antigens
print(decon_results$praOnly)

# Summary of the run
print(decon_results$summary)
```

## epletAUC

The `epletAUC()` function (and its parent, `calculateAUC`) is designed to assess feature reactivity in a dynamic way. Instead of relying on a single MFI cutoff, it calculates the proportion of positive beads for each feature (like an eplet or CREG) across a whole range of cutoffs. By integrating this curve, it produces an Area Under the Curve (AUC) value, which serves as a robust summary of reactivity that is less sensitive to any single arbitrary cutoff.

This function can return either a plot of the positivity curves or a data frame of the calculated AUC values.

**Key Arguments:**

*   `result_file`: Your SAB data, as a data frame or file path.
*   `analysis_type`: The feature you want to analyze, e.g., `"eplet"`, `"creg"`, or `"serology"`.
*   `plot_results`: Set to `TRUE` to get a `ggplot` object, or `FALSE` to get a data frame of AUC values.
*   `feature_filter`: A threshold to remove features that appear on too few beads to be reliable.
*   `percPos_filter`: A threshold to filter out features that never reach a certain level of positivity across all cutoffs.

```{r}
plot_result <- epletAUC(
  result_file   = deepMatchR_example[[1]],
  plot_results  = TRUE,         # Request a plot
  cut_min       = 250,          # Starting cutoff
  cut_max       = 10000,        # Ending cutoff
  cut_step      = 250,          # Step between cutoffs
  palette       = "inferno"     # Color palette for lines
)

plot_result
```

If you prefer a data table with the computed AUC values, simply set `plot_results = FALSE`. The function will then compute and return a data frame:

```{r}
auc_result <- epletAUC(
  result_file   = deepMatchR_example[[1]],
  plot_results  = FALSE,  # Return AUC data instead of a plot
  cut_min       = 250,
  cut_max       = 10000,
  cut_step      = 250
)

auc_result
```

We can use more advanced filtering strategies to extract clinically actionable data. `eplet_filter` allows us to consider eplets that appear at least a certain number of times in your assay. Likewise, `percPos_filter` can be used to threshold the maximumn positivity for each eplet across the cutoffs.

```{r}
epletAUC(
  result_file   = deepMatchR_example[[1]],
  label = FALSE,
  plot_results  = TRUE,
  eplet_filter  = 10,
  percPos_filter = 1.0
)
```

## plotEplets

The `plotEplets()` function offers several ways to visualize eplet-level reactivity from SAB data, allowing you to choose the representation that best suits your question. It simplifies the workflow by handling data loading, merging with eplet definitions, and generating one of three distinct plot types.

**Key Arguments:**

*   `result_file`: Your SAB data, as a data frame or file path.
*   `plot_type`: The visualization to create. Can be `"treemap"`, `"bar"`, or `"AUC"`.
    *   **`"treemap"`**: Good for an overview of the most prominent eplets. The area of each tile is proportional to the eplet's importance.
    *   **`"bar"`**: Best for clearly ranking the top eplets based on the proportion of positive beads at a single MFI `cutoff`.
    *   **`"AUC"`**: Ranks eplets by their reactivity across a *range* of cutoffs, providing a more robust performance metric.
*   `cutoff`: The MFI threshold for defining a positive bead (used for `"treemap"` and `"bar"` plots).
*   `evidence_level`: Filters eplets based on their level of experimental evidence (e.g., `"A1"`, `"A2"`, `"B"`).

### Generating a Treemap

In the treemap visualization, the area of each rectangle represents the product of the number of positive beads and the positive bead percentage for each eplet. This plot is ideal for getting an overall sense of eplet performance across the assay.

```{r}
plotEplets(result_file = deepMatchR_example[[1]],
           plot_type = "treemap",
           cutoff = 2000,
           evidence_level = c("A1", "A2", "B"),
           percPos_filter = 0.4,
           palette = "spectral")
```

### Generating a Barplot

The bar plot ranks eplets by the proportion of positive beads defined by the cutoff. This visualization is useful when you want to focus on the top-performing eplets based on their positivity rate

```{r}
plotEplets(result_file = deepMatchR_example[[1]],
           plot_type = "bar",
           group_by = "loci",
           cutoff = 2000,
           evidence_level = c("A1", "A2", "B"),
           percPos_filter = 0.4,
           top_eplets = 20,
           palette = "spectral")
```

### Generating AUC Barplot

The AUC plot visualizes the normalized area under the curve for eplets across a range of MFI cutoffs. This approach provides insight into the performance consistency of each eplet across different assay thresholds.

```{r}
plotEplets(result_file = deepMatchR_example[[1]],
           plot_type = "AUC",
           percPos_filter = 0.4,
           group_by = "evidence_level",
           cut_min = 250,
           cut_max = 10000,
           cut_step = 250,
           top_eplets = 20,
           palette = "spectral")
```

***

# Conclusions

If you have any questions, comments or suggestions, feel free to visit the [github repository](https://github.com/ncborcherding/deepMatchR).

```{r}
sessionInfo()
```