---
title: "Getting Started with deepMatchR"
author:
- name: Nick Borcherding
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
package: deepMatchR
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Getting Started with deepMatchR}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  collapse = TRUE,
  comment = "#>"
)
library(BiocStyle)

# Check if running on Windows (for skipping Python-dependent sections)
is_windows <- .Platform$OS.type == "windows"
```

# Introduction

Human leukocyte antigen (HLA) molecules are central to both innate and adaptive immune responses. Characterizing HLA antibodies and typing plays a pivotal role in bone marrow and solid organ transplantation.

**deepMatchR** provides a collection of tools that use advanced statistical approaches and deep learning to better characterize HLA immunogenicity for clinical and research applications.

**Key capabilities include:**

- Sequence-level mismatch quantification (amino acid and eplet)
- Peptide-MHC binding prediction
- Visualization of antibody reactivity data
- Integration with standard HLA nomenclature

For more information, visit the [deepMatchR GitHub repository](https://github.com/BorchLab/deepMatchR).

## Installation

```{r install, eval=FALSE}
# Install from Bioconductor (when available)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("deepMatchR")

# Or install the development version from GitHub
# remotes::install_github("BorchLab/deepMatchR")
```

## Quick Start

```{r quickstart}
library(deepMatchR)
library(dplyr)
library(ggplot2)
```

# Working with HLA Data

## Built-in Example Data

**deepMatchR** includes simulated assay data for learning and testing:

1. **Class I SAB** - Single Antigen Bead assay results
2. **Class II SAB** - Single Antigen Bead assay results
3. **PRA** - Panel Reactive Antibody assay results

```{r load-data}
data("deepMatchR_example")
```

### Data Format Requirements

For downstream analysis, your data must include these columns:

| Column | Description | Example |
|--------|-------------|---------|
| `BeadID` | Unique bead identifier | 1, 2, 3... |
| `SpecAbbr` | Abbreviated specificity | A01, B35 |
| `Specificity` | Full specificity string | "A*01:01, B*07:01" |
| `NormalValue` | Measured MFI value | 15234 |

```{r show-data}
# Class I SAB example
head(deepMatchR_example[[1]], 3)

# Class II SAB example
head(deepMatchR_example[[2]], 3)

# PRA example
head(deepMatchR_example[[3]], 3)
```

## Creating HLA Genotypes

The `hlaGeno()` function creates standardized genotype objects used throughout the package.

```{r create-genotypes}
# Define recipient and donor genotypes
recipient <- data.frame(
  A_1 = "A*01:01", A_2 = "A*02:01",
  B_1 = "B*07:02", B_2 = "B*08:01",
  DQA1_1 = "DQA1*02:01", DQA1_2 = "DQA1*05:05",
  DQB1_1 = "DQB1*02:02", DQB1_2 = "DQB1*03:01"
)

donor <- data.frame(
  A_1 = "A*03:01", A_2 = "A*24:02",
  B_1 = "B*44:02", B_2 = "B*51:01",
  DQA1_1 = "DQA1*05:05", DQA1_2 = "DQA1*01:02",
  DQB1_1 = "DQB1*06:02", DQB1_2 = "DQB1*03:01"
)

# Create genotype objects
rgeno <- hlaGeno(recipient)
dgeno <- hlaGeno(donor)
```

# Sequence-Level Analysis

## Retrieving Allele Sequences

The `getAlleleSequence()` function retrieves protein sequences from the IMGT/HLA database.

```{r get-sequence, results='hide'}
# Get the amino acid sequence for A*01:01
a0101_seq <- getAlleleSequence("A*01:01")
```

```{r show-sequence}
# View first 50 amino acids
substr(a0101_seq, 1, 50)
```

## Quantifying Amino Acid Mismatches

The `quantifyMismatch()` function compares two protein sequences and counts differences. You can optionally filter by biophysical properties.

### Basic Usage

```{r mismatch-basic}
seq1 <- "YFAMYGEKVAHTHVDTLYVRYHY"
seq2 <- "YFDMYGEKVAHTHVDTLYVRFHY"

# Count all mismatches
quantifyMismatch(seq1, seq2)
```

### Filtering by Biophysical Properties

**deepMatchR** uses these classification models:

- **Charge:** K/R/H = positive; D/E = negative; others = neutral
- **Polarity:** Nonpolar {A,V,L,I,P,M,F,W,G}; Polar {S,T,N,Q,Y,C,H,K,R,D,E}

```{r mismatch-filters}
# Only charge-changing mismatches
quantifyMismatch(seq1, 
                 seq2, 
                 filter_charge = TRUE)

# Only polarity-changing mismatches
quantifyMismatch(seq1, 
                 seq2, 
                 filter_polarity = TRUE)

# Both charge AND polarity changing
quantifyMismatch(seq1, 
                 seq2, 
                 filter_charge = TRUE, 
                 filter_polarity = TRUE)
```

### Detailed Position-by-Position Analysis

```{r mismatch-detail}
details <- quantifyMismatch(seq1, 
                            seq2, 
                            return = "detail")
head(details)
```

The detail output includes:

- `is_mismatch`: Whether amino acids differ at this position
- `charge_change` / `polarity_change`: Whether that property changed
- `counted`: Whether this position counts toward the final tally

### Function Reference

| Argument | Default | Description |
|----------|---------|-------------|
| `sequence1`, `sequence2` | -- | Two protein sequences (same length) |
| `filter_charge` | `NULL` | Filter by charge change (`TRUE`/`FALSE`/`NULL`) |
| `filter_polarity` | `NULL` | Filter by polarity change (`TRUE`/`FALSE`/`NULL`) |
| `return` | `"count"` | Output type: `"count"` or `"detail"` |
| `na_action` | `"exclude"` | Handle unknown residues: `"exclude"`, `"error"`, `"count"` |

## Visuaizing Sequence Differences

The `plotHLASequences()` function provides comprehensive visualization for comparing
two or more HLA allele sequences. It combines sequence retrieval via `getAlleleSequence()`
with mismatch analysis using `quantifyMismatch()` to create publication-quality alignment
visualizations.

### Basic Usage: Comparing Two Sequences

```{r plot-basic, fig.width=10, fig.height=6}
# Compare two HLA-A alleles
result <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "protein"
)

print(result)
```

The function returns a list containing:

- `plot`: Combined visualization using patchwork
- `mismatch_summary`: Data frame of pairwise mismatch statistics
- `mismatch_details`: Detailed position-by-position comparisons
- `sequences`: The sequences used for comparison
- `plots`: Individual plot components for customization

```{r mismatch-summary}
# View mismatch summary
result$mismatch_summary
```

### Comparing Three or More Sequences

`plotHLASequences()` supports any number of sequences (2 or more):

```{r plot-three, fig.width=10, fig.height=8}
# Compare three DRB1 alleles
result_drb1 <- plotHLASequences(
  alleles = c("DRB1*03:01", "DRB1*04:01", "DRB1*07:01"),
  seq_type = "protein"
)

print(result_drb1)
```

### Using Custom Sequences

You can also provide your own sequences instead of fetching from IMGT/HLA:

```{r plot-custom, fig.width=10, fig.height=5}
# Define custom sequences
custom_seqs <- list(
  "Peptide_A" = "YFAMYGEKVAHTHVDTLYVRYHY",
  "Peptide_B" = "YFDMYGEKVAHTHVDTLYVRFHY",
  "Peptide_C" = "YFAMYGEKVAHTHVDTLYVRFHY"
)

# Compare custom sequences
result_custom <- plotHLASequences(
  alleles = names(custom_seqs),
  sequences = custom_seqs,
  seq_type = "protein",
  show_exons = FALSE  # No exon boundaries for custom sequences
)
print(result_custom)
```

### Applying Biophysical Filters

Filter mismatches by charge or polarity changes:

```{r plot-filters, fig.width=10, fig.height=6, results='hide'}
# Only show charge-changing mismatches
result_charge <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "protein",
  filter_charge = TRUE
)

# Compare filtered vs unfiltered counts
result_charge$mismatch_summary
```

### Focusing on Mismatch Regions

Use `focus_on_mismatches = TRUE` to add a zoomed view:

```{r plot-zoom, fig.width=10, fig.height=8, results='hide'}
result_zoom <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "protein",
  focus_on_mismatches = TRUE,
  context_window = 10  # Show 10 positions around mismatches
)

print(result_zoom)
```

### Nucleotide Sequence Comparison

Compare nucleotide sequences by setting `seq_type = "nucleotide"`:

```{r plot-nuc, fig.width=10, fig.height=6}
result_nuc <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "nucleotide"
)

print(result_nuc)
```

### Color Schemes

Three color schemes are available for protein sequences:

- `"properties"` (default): Groups amino acids by physicochemical properties
- `"hydropathy"`: Colors by hydrophobicity (red = hydrophobic, blue = hydrophilic)
- `"classic"`: Rainbow color scheme

```{r plot-hydropathy, fig.width=10, fig.height=6}
result_hydro <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "protein",
  color_scheme = "hydropathy"
)
print(result_hydro)
```

# Donor-Recipient Matching

## Calculating Mismatch Load

The `calculateMismatchLoad()` function aggregates amino acid mismatches across HLA loci between a donor and recipient.

### Total Mismatch Load

```{r mismatch-load-total}
calculateMismatchLoad(rgeno, 
                      dgeno, 
                      parallel = FALSE)
```

### Per-Locus Breakdown

```{r mismatch-load-locus}
per_locus <- calculateMismatchLoad(
  rgeno, dgeno,
  return = "per_locus",
  parallel = FALSE
)
per_locus
```

### Pairwise Allele Matrix

View mismatches between each recipient-donor allele pair at a specific locus:

```{r mismatch-load-pairwise}
mB <- calculateMismatchLoad(
  rgeno, dgeno,
  return = "pairwise",
  pairwise_locus = "B",
  parallel = FALSE
)
mB
```

### Applying Filters

```{r mismatch-load-filters}
# Charge-changing mismatches only
calculateMismatchLoad(rgeno, 
                      dgeno, 
                      filter_charge = TRUE, 
                      parallel = FALSE)

# Polarity-changing mismatches only
calculateMismatchLoad(rgeno, 
                      dgeno, 
                      filter_polarity = TRUE, 
                      parallel = FALSE)

# Restrict to specific loci
calculateMismatchLoad(rgeno, 
                      dgeno, 
                      loci = "A", 
                      parallel = FALSE)
```

### Function Reference

| Argument | Default | Description |
|----------|---------|-------------|
| `recipient_geno`, `donor_geno` | -- | HLA genotypes from `hlaGeno()` |
| `loci` | `NULL` | Restrict to specific loci (e.g., `c("A","B")`) |
| `filter_charge`, `filter_polarity` | `NULL` | Biophysical filters |
| `return` | `"total"` | `"total"`, `"per_locus"`, or `"pairwise"` |
| `pairwise_locus` | `NULL` | Locus for pairwise matrix |
| `parallel` | `TRUE` | Use parallel processing |

# Eplet Analysis

Eplets are short amino acid sequences that form antibody-binding epitopes on HLA molecules. They are key determinants of HLA immunogenicity.

## Quantifying Eplet Mismatches

The `quantifyEpletMismatch()` function calculates the number of non-shared eplets between two alleles.

```{r eplet-mismatch}
# Basic eplet comparison (A1/A2 evidence level)
quantifyEpletMismatch("A*01:01", 
                      "A*02:01")

# Filter by exposition level
quantifyEpletMismatch("B*07:02", 
                      "B*44:02", 
                      exposition_filter = "High")

# Filter by reactivity confirmation
quantifyEpletMismatch("C*07:01", 
                      "C*06:02",
                      evidence_level = NULL,
                      reactivity_filter = "Confirmed")
```

## Calculating Eplet Load

The `calculateEpletLoad()` function calculates total donor-specific eplets across all loci.

### Total Eplet Load

```{r eplet-load-total}
calculateEpletLoad(rgeno, 
                   dgeno)
```

### Per-Locus Summary

```{r eplet-load-locus}
per <- calculateEpletLoad(rgeno, 
                          dgeno, 
                          return = "per_locus")
per
```

### Pairwise Allele Matrix

```{r eplet-load-pairwise}
mB <- calculateEpletLoad(
  rgeno, dgeno,
  return = "pairwise",
  pairwise_locus = "B"
)
mB
```

### Applying Evidence and Exposition Filters

```{r eplet-load-filters}
# High-exposition eplets only
calculateEpletLoad(rgeno, 
                   dgeno, 
                   exposition_filter = "High")

# Antibody-confirmed eplets only
calculateEpletLoad(rgeno, 
                   dgeno, 
                   reactivity_filter = "Confirmed")

# All evidence levels (no filter)
calculateEpletLoad(rgeno, 
                   dgeno, 
                   evidence_level = NULL)
```

# Peptide Binding Prediction

## MHCnuggets Integration

The `predictMHCnuggets()` function provides an R interface to the MHCnuggets deep learning model for peptide-MHC binding prediction.

> **Note:** This section requires Python dependencies and is skipped on Windows due to path length limitations with the TensorFlow/MHCnuggets installation.

```{r mhcnuggets, eval=!is_windows}
# Define peptides and allele
peptides <- c("SIINFEKL", "LLFGYPVYV")
allele <- "A*02:01"

# Predict binding affinity (IC50)
binding_results <- predictMHCnuggets(
  peptides = peptides,
  allele = allele,
  mhc_class = "I",
  rank_output = TRUE
)

print(binding_results)
```

## Calculating Peptide Binding Load

The `calculatePeptideBindingLoad()` function predicts transplant risk by analyzing peptide-HLA binding between recipient molecules and donor-derived peptides.

**Workflow:**

1. **Input processing** - Standardizes recipient/donor allele formats
2. **Peptide derivation** - Generates mismatched peptides from sequence differences
3. **Binding prediction** - Predicts IC50 using PWM, NetMHCpan, or MHCflurry
4. **Risk calculation** - Aggregates predictions into risk scores

### Total Risk Score

```{r peptide-total}
total_risk <- calculatePeptideBindingLoad(
  recipient = rgeno,
  donor = dgeno,
  return = "total"
)
print(total_risk)
```

### Per-Allele Summary

```{r peptide-summary}
summary_load <- calculatePeptideBindingLoad(
  recipient = rgeno,
  donor = dgeno,
  return = "summary"
)
print(summary_load)
```

### Detailed Peptide-Level Results

```{r peptide-detail}
detailed_load <- calculatePeptideBindingLoad(
  recipient = rgeno,
  donor = dgeno,
  return = "detail"
)
head(detailed_load)
```

# Visualizing Antibody Data

## Antibody Plots with `plotAntibodies()`

The `plotAntibodies()` function creates publication-quality visualizations of SAB and PRA assay data.

### SAB Bar Plot

```{r plot-sab, fig.width=10, fig.height=8}
plotAntibodies(
  result_file = deepMatchR_example[[1]],
  type = "SAB",
  bead_cutoffs = c(2000, 1000, 500, 250),
  add_table = TRUE,
  palette = "spectral",
  highlight_antigen = c("Bw4")
)
```

### PRA Bar Plot

```{r plot-pra, fig.width=10, fig.height=6}
plotAntibodies(
  result_file = deepMatchR_example[[3]],
  type = "PRA",
  class = "I",
  add_table = TRUE,
  palette = "spectral"
)
```

### Time-Series Trend Plot

Track antibody levels over multiple time points:

```{r plot-trend, fig.width=10, fig.height=6}
# Simulate longitudinal data
set.seed(123)
sab_data_list <- list(
  "01/01/2023" = deepMatchR_example[[1]],
  "02/01/2023" = deepMatchR_example[[1]] %>%
    mutate(NormalValue = NormalValue * runif(n(), 0.5, 0.8)),
  "03/01/2023" = deepMatchR_example[[1]] %>%
    mutate(NormalValue = NormalValue * runif(n(), 1, 3))
)

plotAntibodies(
  result_file = sab_data_list,
  type = "SAB",
  plot_trend = TRUE,
  highlight_threshold = 5000,
  vline_dates = c("2023-02-15")
)
```

## Eplet Reactivity Analysis with `epletAUC()`

The `epletAUC()` function calculates the Area Under the Curve (AUC) for eplet reactivity across multiple MFI thresholds. This provides a more robust measure than a single cutoff.

### Visualize Reactivity Curves

```{r eplet-auc-plot, fig.width=8, fig.height=6}
epletAUC(
  result_file = deepMatchR_example[[1]],
  group_by = "evidence",
  evidence_level = c("A1", "A2", "B", "D"),
  plot_results = TRUE,
  cut_min = 250,
  cut_max = 10000,
  cut_step = 250,
  palette = "inferno"
)
```

### Get AUC Values as Data

```{r eplet-auc-data}
auc_result <- epletAUC(
  result_file = deepMatchR_example[[1]],
  plot_results = FALSE,
  cut_min = 250,
  cut_max = 10000,
  cut_step = 250
)
head(auc_result)
```

### Advanced Filtering

```{r eplet-auc-filter, fig.width=8, fig.height=6}
epletAUC(
  result_file = deepMatchR_example[[1]],
  label = FALSE,
  plot_results = TRUE,
  eplet_filter = 10,      # Minimum bead count
  percPos_filter = 1.0    # Minimum positivity threshold
)
```

## Eplet Visualization with `plotEplets()`

The `plotEplets()` function offers three visualization types for eplet reactivity data.

### Treemap View

Ideal for an overview of eplet prominence. Tile area represents importance (positive beads x positivity rate):

```{r plot-treemap, fig.width=10, fig.height=8}
plotEplets(
  result_file = deepMatchR_example[[1]],
  plot_type = "treemap",
  cutoff = 2000,
  evidence_level = c("A1", "A2", "B"),
  percPos_filter = 0.4,
  palette = "spectral"
)
```

### Bar Plot

Ranks eplets by positivity rate at a specific threshold:

```{r plot-bar, fig.width=8, fig.height=6}
plotEplets(
  result_file = deepMatchR_example[[1]],
  plot_type = "bar",
  group_by = "loci",
  cutoff = 2000,
  evidence_level = c("A1", "A2", "B"),
  percPos_filter = 0.4,
  top_eplets = 20,
  palette = "spectral"
)
```

### AUC Bar Plot

Ranks eplets by performance across multiple thresholds:

```{r plot-auc-bar, fig.width=8, fig.height=6}
plotEplets(
 result_file = deepMatchR_example[[1]],
  plot_type = "AUC",
  percPos_filter = 0.4,
  group_by = "evidence_level",
  cut_min = 250,
  cut_max = 10000,
  cut_step = 250,
  top_eplets = 20,
  palette = "spectral"
)
```

# Session Information

```{r sessioninfo}
sessionInfo()
```

# Getting Help

If you have questions, suggestions, or encounter issues:

- **GitHub Issues:** [github.com/BorchLab/deepMatchR/issues](https://github.com/ncborcherding/deepMatchR/issues)
