---
title: "Peptide Binding Prediction"
author:
- name: Nick Borcherding
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
package: deepMatchR
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Peptide Binding Prediction}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  collapse = TRUE,
  comment = "#>"
)
library(BiocStyle)

# Check if running on Windows (for skipping Python-dependent sections)
is_windows <- .Platform$OS.type == "windows"
```

# Introduction

Peptide-MHC binding prediction is central to understanding T cell-mediated immunity. In the context of transplantation, predicting how recipient HLA molecules might present donor-derived peptides helps assess immunological risk.

This article covers:

- Using MHCnuggets for peptide binding prediction
- Calculating peptide binding load for donor-recipient pairs
- Interpreting binding predictions

# Loading the Package

```{r load-package}
library(deepMatchR)
library(dplyr)
```

# Background

## Peptide-MHC Binding

MHC molecules present peptides to T cells. The binding affinity between a peptide and an MHC molecule determines:

- **Presentation likelihood:** Strong binders are more likely to be presented
- **T cell activation potential:** Better presentation leads to stronger T cell responses
- **Immunogenicity:** Peptides that bind well and differ from self are immunogenic

## Binding Thresholds

Binding affinity is typically measured as IC50 (half-maximal inhibitory concentration):

| IC50 (nM) | Classification |
|-----------|----------------|
| < 50 | Strong binder |
| 50-500 | Intermediate binder |
| 500-5000 | Weak binder |
| > 5000 | Non-binder |

# MHCnuggets Prediction

## Overview

`predictMHCnuggets()` provides an R interface to MHCnuggets, a deep learning model for peptide-MHC binding prediction.

> **Note:** MHCnuggets requires Python dependencies via basilisk. This functionality is not available on Windows due to path length limitations.

## Basic Usage

```{r mhcnuggets, eval=!is_windows}
# Define peptides and allele
peptides <- c("SIINFEKL", "LLFGYPVYV", "GILGFVFTL")
allele <- "A*02:01"

# Predict binding affinity
binding_results <- predictMHCnuggets(
  peptides = peptides,
  allele = allele,
  mhc_class = "I"
)

print(binding_results)
```

## Interpretation

The output includes:

- **peptide:** Input peptide sequence
- **allele:** HLA allele tested
- **ic50:** Predicted binding affinity in nM
- **rank:** Percentile rank (if `rank_output = TRUE`)

# Peptide Binding Load

## Concept

The `calculatePeptideBindingLoad()` function estimates transplant risk by:

1. Identifying mismatched peptides from donor sequences
2. Predicting binding to recipient HLA molecules
3. Aggregating predictions into a risk score

## Creating Genotypes

```{r create-genotypes}
recipient <- data.frame(
  A_1 = "A*02:01", A_2 = "A*03:01",
  B_1 = "B*07:02", B_2 = "B*08:01"
)

donor <- data.frame(
  A_1 = "A*01:01", A_2 = "A*24:02",
  B_1 = "B*44:02", B_2 = "B*51:01"
)

rgeno <- hlaGeno(recipient)
dgeno <- hlaGeno(donor)
```

## Total Risk Score

```{r peptide-total}
total_risk <- calculatePeptideBindingLoad(
  recipient = rgeno,
  donor = dgeno,
  return = "total"
)
print(total_risk)
```

## Per-Allele Summary

```{r peptide-summary}
summary_load <- calculatePeptideBindingLoad(
  recipient = rgeno,
  donor = dgeno,
  return = "summary"
)
print(summary_load)
```

## Detailed Peptide Results

```{r peptide-detail}
detailed_load <- calculatePeptideBindingLoad(
  recipient = rgeno,
  donor = dgeno,
  return = "detail"
)
head(detailed_load)
```

# Backends

## Available Backends

`calculatePeptideBindingLoad()` supports multiple prediction backends:

| Backend | Description | Requirements |
|---------|-------------|--------------|
| `pwm` | Position Weight Matrix (default) | R only |
| `mhcnuggets` | Deep learning model | Python |
| `netmhcpan` | External tool | NetMHCpan installed |
| `mhcflurry` | Deep learning model | Python |

## Using PWM (Default)

The PWM backend uses pre-computed position-specific scoring matrices based on HLA supertypes:

```{r pwm-backend}
result_pwm <- calculatePeptideBindingLoad(
  recipient = rgeno,
  donor = dgeno,
  backend = "pwm",
  return = "summary"
)
print(result_pwm)
```

## Customizing Thresholds

```{r custom-thresholds}
# Adjust binding thresholds
result_custom <- calculatePeptideBindingLoad(
  recipient = rgeno,
  donor = dgeno,
  binding_threshold = 100,   # Strong binder threshold
  weak_threshold = 1000,     # Weak binder threshold
  return = "summary"
)
print(result_custom)
```

# Clinical Applications

## Transplant Risk Assessment

Peptide binding load can complement eplet-based analysis:

1. **Eplet load:** Predicts antibody-mediated rejection risk
2. **Peptide binding load:** Predicts T cell-mediated rejection risk

## Integration with Other Metrics

Consider combining peptide binding predictions with:

- HLA mismatch count
- Eplet mismatch load
- Pre-existing antibody profiles
- Clinical risk factors

# Function Reference

## `predictMHCnuggets()`

| Argument | Default | Description |
|----------|---------|-------------|
| `peptides` | -- | Character vector of peptide sequences |
| `allele` | -- | HLA allele (e.g., "A*02:01") |
| `mhc_class` | `"I"` | MHC class: "I" or "II" |
| `rank_output` | `FALSE` | Include percentile rank |

## `calculatePeptideBindingLoad()`

| Argument | Default | Description |
|----------|---------|-------------|
| `recipient` | -- | Recipient genotype or allele vector |
| `donor` | -- | Donor peptides or genotype |
| `backend` | `"pwm"` | Prediction backend |
| `peptide_length` | `9L` | Peptide length for analysis |
| `binding_threshold` | `500` | IC50 threshold for strong binding |
| `weak_threshold` | `5000` | IC50 threshold for weak binding |
| `return` | `"total"` | Output type: "total", "summary", "detail" |
| `aggregate_method` | `"sum"` | Aggregation: "sum", "max", "mean" |

# Session Information

```{r sessioninfo}
sessionInfo()
```
