---
title: "Sequence & Mismatch Analysis"
author:
- name: Nick Borcherding
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
package: deepMatchR
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Sequence & Mismatch Analysis}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  collapse = TRUE,
 comment = "#>"
)
library(BiocStyle)
```

# Introduction

This article covers sequence-level HLA analysis in **deepMatchR**, including:

- Retrieving allele sequences from IMGT/HLA
- Quantifying amino acid mismatches between sequences
- Visualizing and comparing multiple HLA sequences
- Calculating mismatch load between donor-recipient pairs

These tools provide the foundation for understanding immunogenicity at the molecular level.

# Loading the Package

```{r load-package}
library(deepMatchR)
library(dplyr)
```

# Retrieving HLA Sequences

## Basic Usage

The `getAlleleSequence()` function retrieves protein sequences from the IMGT/HLA database.

```{r get-sequence}
# Get the amino acid sequence for A*01:01
a0101_seq <- getAlleleSequence("A*01:01")

# View first 60 amino acids
substr(a0101_seq, 1, 60)

# Full sequence length
nchar(a0101_seq)
```

## Batch Retrieval

For multiple alleles, use `batchGetSequences()` for efficient parallel retrieval:
```{r batch-sequences, results='hide'}
alleles <- c("A*01:01", "A*02:01", "B*07:02")
sequences <- batchGetSequences(alleles)
```

```{r show-batch}
# View sequence lengths
sapply(sequences, nchar)
```

# Quantifying Amino Acid Mismatches

## Basic Mismatch Counting

The `quantifyMismatch()` function compares two protein sequences and counts differences:

```{r mismatch-basic}
seq1 <- "YFAMYGEKVAHTHVDTLYVRYHY"
seq2 <- "YFDMYGEKVAHTHVDTLYVRFHY"

# Count all mismatches
quantifyMismatch(seq1, seq2)
```

## Biophysical Property Filters

You can filter mismatches by biophysical properties to focus on functionally significant changes:

### Charge Classification

- **Positive:** K, R, H
- **Negative:** D, E
- **Neutral:** All others

### Polarity Classification

- **Nonpolar:** A, V, L, I, P, M, F, W, G
- **Polar:** S, T, N, Q, Y, C, H, K, R, D, E

```{r mismatch-filters}
# Only charge-changing mismatches
quantifyMismatch(seq1, seq2, filter_charge = TRUE)

# Only polarity-changing mismatches
quantifyMismatch(seq1, seq2, filter_polarity = TRUE)

# Both charge AND polarity changing
quantifyMismatch(seq1, seq2, filter_charge = TRUE, filter_polarity = TRUE)
```

## Detailed Position Analysis

Get position-by-position breakdown:

```{r mismatch-detail}
details <- quantifyMismatch(seq1, seq2, return = "detail")
head(details)
```

The detail output includes:

| Column | Description |
|--------|-------------|
| `position` | Position in the sequence |
| `aa1` / `aa2` | Amino acid in each sequence |
| `is_mismatch` | Whether amino acids differ |
| `charge_change` | Whether charge class changed |
| `polarity_change` | Whether polarity changed |
| `counted` | Whether position counts toward total |

## Handling Unknown Residues

The `na_action` parameter controls how unknown residues (X, -, etc.) are handled:

```{r na-action}
seq_with_unknown <- "YFAMYGEKVAHTHVDTLYVRXHY"

# Exclude unknown positions (default)
quantifyMismatch(seq1, seq_with_unknown, na_action = "exclude")

# Count unknowns as mismatches
quantifyMismatch(seq1, seq_with_unknown, na_action = "count")
```

# Visualizing HLA Sequence Alignments

## Introduction to plotHLASequences

The `plotHLASequences()` function provides comprehensive visualization for comparing
two or more HLA allele sequences. It combines sequence retrieval via `getAlleleSequence()`
with mismatch analysis using `quantifyMismatch()` to create publication-quality alignment
visualizations.

## Basic Usage: Comparing Two Sequences

```{r plot-basic, fig.width=10, fig.height=6}
# Compare two HLA-A alleles
result <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "protein"
)

# Display the plot
print(result)
```

The function returns a list containing:

- `plot`: Combined visualization using patchwork
- `mismatch_summary`: Data frame of pairwise mismatch statistics
- `mismatch_details`: Detailed position-by-position comparisons
- `sequences`: The sequences used for comparison
- `plots`: Individual plot components for customization

```{r mismatch-summary}
# View mismatch summary
result$mismatch_summary
```

## Comparing Three or More Sequences

`plotHLASequences()` supports any number of sequences (2 or more):

```{r plot-three, fig.width=10, fig.height=8}
# Compare three DRB1 alleles
result_drb1 <- plotHLASequences(
  alleles = c("DRB1*03:01", "DRB1*04:01", "DRB1*07:01"),
  seq_type = "protein"
)

print(result_drb1)
```

## Using Custom Sequences

You can also provide your own sequences instead of fetching from IMGT/HLA:

```{r plot-custom, fig.width=10, fig.height=5}
# Define custom sequences
custom_seqs <- list(
  "Peptide_A" = "YFAMYGEKVAHTHVDTLYVRYHY",
  "Peptide_B" = "YFDMYGEKVAHTHVDTLYVRFHY",
  "Peptide_C" = "YFAMYGEKVAHTHVDTLYVRFHY"
)

# Compare custom sequences
result_custom <- plotHLASequences(
  alleles = names(custom_seqs),
  sequences = custom_seqs,
  seq_type = "protein",
  show_exons = FALSE  # No exon boundaries for custom sequences
)
print(result_custom)
```

## Applying Biophysical Filters

Filter mismatches by charge or polarity changes:

```{r plot-filters, fig.width=10, fig.height=6}
# Only show charge-changing mismatches
result_charge <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "protein",
  filter_charge = TRUE
)

# Compare filtered vs unfiltered counts
result_charge$mismatch_summary
```

## Focusing on Mismatch Regions

Use `focus_on_mismatches = TRUE` to add a zoomed view:

```{r plot-zoom, fig.width=10, fig.height=8}
result_zoom <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "protein",
  focus_on_mismatches = TRUE,
  context_window = 10  # Show 10 positions around mismatches
)

print(result_zoom)
```

## Nucleotide Sequence Comparison

Compare nucleotide sequences by setting `seq_type = "nucleotide"`:

```{r plot-nuc, fig.width=10, fig.height=6}
result_nuc <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "nucleotide"
)

print(result_nuc)
```

## Color Schemes

Three color schemes are available for protein sequences:

- `"properties"` (default): Groups amino acids by physicochemical properties
- `"hydropathy"`: Colors by hydrophobicity (red = hydrophobic, blue = hydrophilic)
- `"classic"`: Rainbow color scheme

```{r plot-hydropathy, fig.width=10, fig.height=6}
result_hydro <- plotHLASequences(
  alleles = c("A*01:01", "A*02:01"),
  seq_type = "protein",
  color_scheme = "hydropathy"
)

print(result_hydro)
```

## Accessing Detailed Results

The returned object contains detailed mismatch information:

```{r detailed-results}
# Get detailed mismatch data for the first comparison
details <- result$mismatch_details[[1]]

# Filter to only mismatch positions
mismatches_only <- details[details$is_mismatch, ]
head(mismatches_only[, c("alignment_position", "ref", "alt",
                         "charge_change", "polarity_change")])
```

# Donor-Recipient Mismatch Load

## Creating HLA Genotypes

First, create genotype objects using `hlaGeno()`:

```{r create-genotypes}
recipient <- data.frame(
  A_1 = "A*01:01", A_2 = "A*02:01",
  B_1 = "B*07:02", B_2 = "B*08:01",
  DRB1_1 = "DRB1*03:01", DRB1_2 = "DRB1*04:01"
)

donor <- data.frame(
  A_1 = "A*03:01", A_2 = "A*24:02",
  B_1 = "B*44:02", B_2 = "B*51:01",
  DRB1_1 = "DRB1*07:01", DRB1_2 = "DRB1*11:01"
)

rgeno <- hlaGeno(recipient)
dgeno <- hlaGeno(donor)
```

## Calculating Total Mismatch Load

```{r mismatch-load-total}
calculateMismatchLoad(rgeno, dgeno, parallel = FALSE)
```

## Per-Locus Breakdown

```{r mismatch-load-locus}
per_locus <- calculateMismatchLoad(
  rgeno, dgeno,
  return = "per_locus",
  parallel = FALSE
)
per_locus
```

## Pairwise Allele Matrix

View mismatches between each recipient-donor allele pair:

```{r mismatch-load-pairwise}
mB <- calculateMismatchLoad(
  rgeno, dgeno,
  return = "pairwise",
  pairwise_locus = "B",
  parallel = FALSE
)
mB
```

## Applying Filters

```{r mismatch-load-filters}
# Charge-changing mismatches only
calculateMismatchLoad(rgeno, dgeno, filter_charge = TRUE, parallel = FALSE)

# Polarity-changing mismatches only
calculateMismatchLoad(rgeno, dgeno, filter_polarity = TRUE, parallel = FALSE)

# Restrict to specific loci
calculateMismatchLoad(rgeno, dgeno, loci = c("A", "B"), parallel = FALSE)
```

# Function Reference

## `quantifyMismatch()`

| Argument | Default | Description |
|----------|---------|-------------|
| `sequence1`, `sequence2` | -- | Two protein sequences (same length) |
| `filter_charge` | `NULL` | Filter by charge change |
| `filter_polarity` | `NULL` | Filter by polarity change |
| `return` | `"count"` | Output type: `"count"` or `"detail"` |
| `na_action` | `"exclude"` | Handle unknown residues: `"exclude"`, `"error"`, `"count"` |

## `calculateMismatchLoad()`

| Argument | Default | Description |
|----------|---------|-------------|
| `recipient_geno`, `donor_geno` | -- | HLA genotypes from `hlaGeno()` |
| `loci` | `NULL` | Restrict to specific loci |
| `filter_charge`, `filter_polarity` | `NULL` | Biophysical filters |
| `return` | `"total"` | `"total"`, `"per_locus"`, or `"pairwise"` |
| `pairwise_locus` | `NULL` | Locus for pairwise matrix |
| `parallel` | `TRUE` | Use parallel processing |

## `plotHLASequences()`

| Argument | Default | Description |
|----------|---------|-------------|
| `alleles` | -- | Character vector of HLA allele names (2 or more) |
| `sequences` | `NULL` | Optional named list of sequences (overrides IMGT/HLA fetch) |
| `locus` | `NULL` | HLA locus for exon boundaries (auto-inferred if NULL) |
| `seq_type` | `"protein"` | Sequence type: `"protein"` or `"nucleotide"` |
| `focus_on_mismatches` | `FALSE` | Add zoomed view of mismatch regions |
| `context_window` | `20` | Positions to show around mismatches in zoom view |
| `show_exons` | `TRUE` | Display exon/domain boundaries |
| `color_scheme` | `"properties"` | Color scheme: `"properties"`, `"hydropathy"`, or `"classic"` |
| `reference_idx` | `1` | Index of reference sequence for comparisons |
| `filter_charge` | `NULL` | Filter by charge change (passed to `quantifyMismatch()`) |
| `filter_polarity` | `NULL` | Filter by polarity change (passed to `quantifyMismatch()`) |
| `use_cache` | `TRUE` | Use caching for sequence retrieval |
| `verbose` | `TRUE` | Print progress messages |

# Session Information

```{r sessioninfo}
sessionInfo()
```
